---
title: "Buffered Panel Threshold Regression with bufferedThresholdPanel"
author: "Faycal Hamdi, Said Souam, Messaoud Zouikri"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Buffered Panel Threshold Regression with bufferedThresholdPanel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 3.5
)
```

## Overview

The **bufferedThresholdPanel** package implements the Buffered Panel Threshold
Data (BTPD) model of @Belarbi2021 for panel data with individual fixed effects.
It extends the classical Panel Threshold Regression (PTR) of @Hansen1999 by
replacing a single threshold parameter with a *buffer zone* $(r_L, r_U]$.
When the threshold variable falls inside this zone, the regime indicator keeps
its previous value, producing a hysteresis-driven (path-dependent) transition
between regimes.

The package supports:

- **Two-regime models** — one buffer zone, slope vectors $\beta_1$ and $\beta_2$
- **Three-regime models** — two buffer zones, slope vectors $\beta_1$, $\beta_2$, $\beta_3$
- **Classical PTR** of @Hansen1999 as a special case ($r_L = r_U$)
- **Sequential bootstrap tests** for the number of regimes ($F_{1,2}$ and $F_{2,3}$)
- **Parallel bootstrap** confidence intervals
- **Publication-ready tables** in AER and Journal of Econometrics styles

---

## The Model

### Two-regime BTPD

For a balanced panel $\{Y_{it}, q_{it}, X_{it} : 1 \le i \le N, 1 \le t \le T\}$
with individual fixed effects $\mu_i$:

$$Y_{it} = \mu_i + X_{it}\beta_1 \mathbf{1}(R_{it}=1)
         + X_{it}\beta_2 \mathbf{1}(R_{it}=2) + \varepsilon_{it}$$

The regime indicator follows the hysteresis rule:

$$R_{it} = \begin{cases}
1           & q_{it} \le r_L \\
R_{i,t-1}   & r_L < q_{it} \le r_U \quad \text{(buffer zone: inherit past regime)} \\
2           & q_{it} > r_U
\end{cases}$$

This path-dependent mechanism — formalised in the time-series context by
@Li2015 — means that two units with identical $q_{it}$ values may be assigned
to different regimes depending on their history.

### Three-regime BTPD

With two buffer zones $(r_{L,1}, r_{U,1}]$ and $(r_{L,2}, r_{U,2}]$ satisfying
$r_{L,1} \le r_{U,1} < r_{L,2} \le r_{U,2}$, the outcome equation becomes:

$$Y_{it} = \mu_i + X_{it}\beta_1 \mathbf{1}(R_{it}=1)
         + X_{it}\beta_2 \mathbf{1}(R_{it}=2)
         + X_{it}\beta_3 \mathbf{1}(R_{it}=3) + \varepsilon_{it}$$

The regime indicator is determined by five regions of the threshold variable:

| Region | Assignment rule |
|---|---|
| $q_{it} \le r_{L,1}$ | Regime 1 (unambiguous) |
| $r_{L,1} < q_{it} \le r_{U,1}$ | Buffer zone 1: stay in 1 if $R_{i,t-1}=1$, else go to 2 |
| $r_{U,1} < q_{it} \le r_{L,2}$ | Regime 2 (unambiguous) |
| $r_{L,2} < q_{it} \le r_{U,2}$ | Buffer zone 2: stay in 2 if $R_{i,t-1} \le 2$, else stay in 3 |
| $q_{it} > r_{U,2}$ | Regime 3 (unambiguous) |

### Estimation

Fixed effects $\mu_i$ are removed by within-group demeaning.  For a given set
of buffer-zone parameters $\gamma = (r_{L,1}, r_{U,1}, r_{L,2}, r_{U,2})$, the
regime indicators are deterministic and the model reduces to OLS applied
separately on each regime sub-sample (*concentrated OLS*).  The optimal
$\hat\gamma$ is the candidate that minimises the concentrated sum of squared
residuals over a candidate grid.  Heteroskedasticity-robust standard errors
(HC0–HC3) are computed via the sandwich estimator.

### Sequential tests

Because the threshold parameter is unidentified under the null hypothesis of
linearity, the test statistic does not follow a standard asymptotic distribution
— a problem first analysed by @Davies1987.  Bootstrap p-values are therefore
used throughout.

1. **$F_{1,2}$**: Tests linearity against a two-regime model.
   Bootstrap draws residuals from the *linear* null.
2. **$F_{2,3}$**: Tests the two-regime model against a three-regime model.
   Bootstrap draws residuals from the *two-regime* null — the key distinction
   from $F_{1,2}$ (see @Belarbi2021, Section 2.3).

---

## Installation

```{r install, eval=FALSE}
# From CRAN (once published)
install.packages("bufferedThresholdPanel")

# Development version
remotes::install_github("MessaoudZouikri/bufferedThresholdPanel")
```

```{r load}
library(bufferedThresholdPanel)
```

---

## Quick illustration with simulated data

The examples below use a small simulated panel (N = 30 units, T = 10 periods,
300 observations) so they run quickly.
Section [Replicating Hamdi et al. (2025)](#full) below shows the exact code
for the full `panel_data` application.

```{r simulate}
set.seed(42)
N  <- 30   # cross-sectional units
TT <- 10   # time periods

df_sim <- data.frame(
  id   = rep(seq_len(N), each = TT),
  time = rep(seq_len(TT), N),
  x1   = rnorm(N * TT),
  x2   = rnorm(N * TT),
  q    = rnorm(N * TT)           # threshold variable
)

# True DGP: two regimes separated at q = 0.4
#   Regime 1 (q <= 0.4):  y = 1.5*x1 - 0.5*x2 + e
#   Regime 2 (q >  0.4):  y = -0.5*x1 + 1.0*x2 + e
df_sim$y <- with(df_sim,
  ifelse(q <= 0.4,
         1.5 * x1 - 0.5 * x2,
        -0.5 * x1 + 1.0 * x2) + rnorm(N * TT, 0, 0.5)
)
```

### Estimate a two-regime BTPD model

```{r fit2}
fit2 <- bptr(
  y ~ x1 + x2,
  data     = df_sim,
  id       = "id",
  time     = "time",
  q        = "q",
  n_thresh = 1,          # 1 threshold → 2 regimes
  buffer   = TRUE,       # BTPD (hysteresis) transitions
  se_type  = "HC3"
)
print(fit2)
```

```{r fit2_summary}
summary(fit2)
```

### S3 extractors

```{r extractors}
# Coefficient matrix (variables × regimes)
round(coef(fit2), 4)

# Number of regimes in the fitted object
fit2$n_regimes          # 2 for n_thresh=1; 3 for n_thresh=2

# Number of observations per regime
fit2$n_obs_regime

# Within R-squared
round(1 - fit2$ssr / fit2$tss_within, 4)

# Threshold / buffer-zone boundaries  [rL, rU]
round(fit2$thresholds, 4)

# Variance-covariance matrix (block-diagonal by regime)
round(vcov(fit2), 6)

# In-sample prediction (identical to fitted values)
head(round(predict(fit2), 4))
```

### Estimate a three-regime BTPD model

```{r fit3}
fit3 <- bptr(
  y ~ x1 + x2,
  data        = df_sim,
  id          = "id",
  time        = "time",
  q           = "q",
  n_thresh    = 2,        # 2 thresholds → 3 regimes
  buffer      = TRUE,
  grid_size_3 = 10,       # ⚠ illustration only — use >= 50 for real analyses
  se_type     = "HC3"
)
summary(fit3)
```

```{r fit3_zones}
# Confirm 3 regimes are stored
fit3$n_regimes                               # should be 3
cat("Buffer zone 1 : [", round(fit3$thresholds[1], 4),
    ",", round(fit3$thresholds[2], 4), "]\n")
cat("Buffer zone 2 : [", round(fit3$thresholds[3], 4),
    ",", round(fit3$thresholds[4], 4), "]\n")
cat("Regime sizes  :", paste(fit3$n_obs_regime, collapse = " / "), "\n")
```

### Classical PTR for comparison

```{r fit_ptr}
fit_ptr <- bptr(
  y ~ x1 + x2,
  data     = df_sim,
  id       = "id",
  time     = "time",
  q        = "q",
  n_thresh = 1,
  buffer   = FALSE        # abrupt transition (Hansen 1999)
)
cat("PTR threshold :", round(fit_ptr$thresholds, 4), "\n")
cat("BTPD zone     : [", round(fit2$thresholds, 4), "]\n")
```

---

## Tidy output (broom integration)

```{r tidy}
library(broom)

# Coefficient table with 95% confidence intervals
tidy(fit2, conf.int = TRUE)
```

```{r glance}
# Model-level fit statistics
glance(fit2)
```

```{r threshold_tidy}
# Threshold / buffer-zone estimates
threshold_tidy(fit2)
```

```{r augment}
# Add fitted values and regime classification to the data frame
aug <- augment(fit2)
head(aug[, c("id", "time", "q", "y", ".fitted", ".resid", ".regime")])
```

---

## Diagnostic plots

```{r plot1, fig.height=3.5}
# Residuals vs fitted, coloured by regime
plot(fit2, which = 1)
```

```{r plot2, fig.height=3.5}
# Threshold variable with buffer-zone boundaries
plot(fit2, which = 2)
```

---

## Publication-ready tables

```{r kable}
# knitr::kable for R Markdown documents
bptr_kable(fit2, digits = 3)
```

```{r table_aer, eval=FALSE}
# American Economic Review style (requires gt)
bptr_table(fit2, style = "AER", stars = TRUE,
           title = "Two-Regime BTPD Estimation")

# LaTeX export for inclusion in a paper
bptr_latex(fit2, file = "table_btpd.tex")
```

---

## Sequential tests (bootstrap — not run in vignette)

The sequential tests require many bootstrap replications and take several
minutes on a full panel dataset.  The code below shows the exact calls;
run them interactively in an R session.

**Three separate functions** implement the sequential strategy of @Belarbi2021,
Section 2.3:

- `bptr_test()` — $F_{1,2}$: tests linearity vs a 2-regime model.
  Returns an object of class `"bptr_test"`.
- `bptr_test_23()` — $F_{2,3}$: tests the 2-regime model vs a 3-regime model.
  Takes a *fitted* `bptr` object (not a formula) as its first argument, and
  returns an object of class `"bptr_test23"`.
- `bptr_test_seq()` — runs both tests in sequence, prints step-by-step
  progress messages to the console via `cat()`, and selects the preferred
  number of regimes automatically.

Both BTPD (`buffer = TRUE`) and classical PTR (`buffer = FALSE`) tests are
supported.

```{r seq_test, eval=FALSE}
# ── F1,2: linearity vs 2-regime BTPD ──────────────────────────────────────
test12 <- bptr_test(
  y ~ x1 + x2, data = df_sim,
  id = "id", time = "time", q = "q",
  buffer = TRUE, n_boot = 499, seed = 42
)
print(test12)      # class "bptr_test"

# ── F1,2 for classical PTR (buffer = FALSE) ────────────────────────────────
test12_ptr <- bptr_test(
  y ~ x1 + x2, data = df_sim,
  id = "id", time = "time", q = "q",
  buffer = FALSE, n_boot = 499, seed = 42
)
print(test12_ptr)

# ── F2,3: 2-regime vs 3-regime (takes a fitted bptr object as H0) ──────────
# Note: first argument is the fitted 2-regime model, NOT a formula
test23 <- bptr_test_23(fit2, n_boot = 499, grid_size_3 = 50, seed = 42)
print(test23)      # class "bptr_test23" — different print method from test12

# ── Complete sequential procedure (F1,2 → F2,3 → model selection) ──────────
# bptr_test_seq() prints progress via cat() as each step completes
result_seq <- bptr_test_seq(
  y ~ x1 + x2, data = df_sim,
  id = "id", time = "time", q = "q",
  buffer = TRUE, n_boot = 499, alpha = 0.10, seed = 42
)
print(result_seq)
# $n_regimes_selected reports 1, 2, or 3
# $final_model is the recommended fitted bptr object
```

---

## Bootstrap confidence intervals (not run in vignette)

`bptr_bootstrap()` computes percentile confidence intervals for all threshold
and slope parameters via *within-unit residual resampling*: for each bootstrap
replication, residuals are resampled independently within each panel unit
(preserving cross-sectional dependence), a pseudo-outcome
$y^* = \hat{y} + e^*$ is formed, and the model is re-estimated.
Intervals are the empirical 2.5 % and 97.5 % quantiles of the bootstrap
distribution.  Parallel computation is supported via the `workers` argument.

```{r bootstrap_ci, eval=FALSE}
# Residual bootstrap for the 2-regime BTPD model (parallel on all cores - 1)
boot <- bptr_bootstrap(fit2, n_boot = 499, seed = 42)
print(boot)
# $ci_lower / $ci_upper: named vectors covering gamma1, gamma2,
#   regime1_x1, regime1_x2, regime2_x1, regime2_x2
```

---

## Replicating Hamdi et al. (2025) {#full}

The `panel_data` dataset ships with the package: 92 countries, 2002–2016,
1 380 observations, no missing values.  The code below replicates the
analysis of @Hamdi2025.  Run it interactively — the sequential test and
bootstrap steps take several minutes.

```{r full_data_overview}
data(panel_data)

cat("Countries :", length(unique(panel_data$country)), "\n")
cat("Years     :", min(panel_data$year), "–", max(panel_data$year), "\n")
cat("Obs       :", nrow(panel_data), "\n")
cat("Columns   :", ncol(panel_data), "\n")

# Both threshold variables and the dependent variable
summary(panel_data[, c("growthRate", "oilRentGDP", "rle", "eci")])
```

### Estimation (runs quickly)

```{r full_fit2}
# Model I: oilRentGDP as threshold, rle as predictor
fit2_I <- bptr(
  growthRate ~ rle + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data     = panel_data,
  id       = "countryId",
  time     = "year",
  q        = "oilRentGDP",
  n_thresh = 1,
  buffer   = TRUE,
  se_type  = "HC3"
)
print(fit2_I)

# Model II: rle as threshold, oilRentGDP as predictor
fit2_II <- bptr(
  growthRate ~ oilRentGDP + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data     = panel_data,
  id       = "countryId",
  time     = "year",
  q        = "rle",
  n_thresh = 1,
  buffer   = TRUE,
  se_type  = "HC3"
)
print(fit2_II)
```

### Sequential test and 3-regime model (slow — run interactively)

```{r full_seq, eval=FALSE}
# Model I: oilRentGDP as threshold, rle as predictor
seq_I <- bptr_test_seq(
  growthRate ~ rle + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data        = panel_data,
  id          = "countryId",
  time        = "year",
  q           = "oilRentGDP",
  buffer      = TRUE,
  n_boot      = 499,
  grid_size_3 = 50,
  alpha       = 0.10,
  seed        = 2025
)
print(seq_I)

# Model II: rle as threshold, oilRentGDP as predictor
seq_II <- bptr_test_seq(
  growthRate ~ oilRentGDP + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data        = panel_data,
  id          = "countryId",
  time        = "year",
  q           = "rle",
  buffer      = TRUE,
  n_boot      = 499,
  grid_size_3 = 50,
  alpha       = 0.10,
  seed        = 2025
)
print(seq_II)

# Fine-grid three-regime estimation for the selected model(s)
fit3_I <- bptr(
  growthRate ~ rle + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data        = panel_data,
  id          = "countryId", time = "year",
  q           = "oilRentGDP",
  n_thresh    = 2, buffer = TRUE, grid_size_3 = 80, se_type = "HC3"
)
summary(fit3_I)

fit3_II <- bptr(
  growthRate ~ oilRentGDP + eci + initialGDP + fdiGDP + capFormGDP +
               inflation + popGrowth + indVAGDP + tradeOpenness,
  data        = panel_data,
  id          = "countryId", time = "year",
  q           = "rle",
  n_thresh    = 2, buffer = TRUE, grid_size_3 = 80, se_type = "HC3"
)
summary(fit3_II)

# Bootstrap confidence intervals
boot_I  <- bptr_bootstrap(fit2_I,  n_boot = 499, seed = 2025)
boot_II <- bptr_bootstrap(fit2_II, n_boot = 499, seed = 2025)
print(boot_I); print(boot_II)
```

---

## Grid resolution guide

| Model | Buffer? | Grid parameter | SSR evaluations (approx.) |
|---|---|---|---|
| 2-regime PTR | No | `grid_size = 300` | 300 |
| 2-regime BTPD | Yes | `grid_size = 300` | 45 000 |
| 3-regime PTR | No | `grid_size = 300` | 45 000 |
| 3-regime BTPD | Yes | `grid_size_3 = 50` | 260 000 |

Use smaller values during exploration (`grid_size = 100`, `grid_size_3 = 20`)
and increase to 300 / 50 (or higher) for final published results.

## Citation

If you use this package in published research, please cite the package and
the two underlying methodology papers.  Running the following command inside
R will display the full citation details:

```{r citation, eval=FALSE}
citation("bufferedThresholdPanel")
```

The three entries returned are:

1. **The R package itself** — Hamdi, Souam, and Zouikri (2025), v0.1.0
2. **The empirical application** — Hamdi, Souam, and Zouikri (2025,
   *The World Economy*) — the paper that introduced the two-model framework
   (Model I: `oilRentGDP` as threshold; Model II: `rle` as threshold) and
   the `panel_data` dataset included with this package
3. **The methodology paper** — Belarbi, Hamdi, Khalfi, and Souam (2021,
   *Economic Modelling*) — the paper that formalised the BTPD model and
   the sequential $F_{1,2}$ / $F_{2,3}$ bootstrap tests

---

## References
